---
title: "Data visualization exercise: solutions"
date: 2026-09-01
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  message: false
  warning: false
---

This document includes only one possible solution. Yours may vary.

## Setup (given) {.unnumbered}

```{r}
#| label: setup

library(tidyverse)

counties = left_join(
  read_csv("https://princeton-ddss.github.io/r-bootcamp/files/county_data.csv"),
  read_csv("https://princeton-ddss.github.io/r-bootcamp/files/county_population.csv"),
  join_by(GEOID == fips)
)
```

# Part 1: the county data

## Rent and income distributions by group

```{r}
#| label: task-1

library(scales)

## 12 NA values. Fine.

counties |>
  filter(!is.na(region)) |>
  pivot_longer(c(income, rent)) |>
  mutate(name = str_to_sentence(name)) |>
  ggplot(aes(x = value, color = region)) +
  geom_density(linewidth = 1) +
  facet_wrap(~ name, scales = "free") +
  labs(x = NULL, y = "Density", color = "Region") +
  theme_bw(base_size = 20) +
  theme(
    legend.position = 'bottom',
    axis.text.y = element_blank(),
    strip.background = element_rect(
      fill = 'white',
      color = "black",
      linewidth = 0
      ),
    ) +
  scale_x_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  expand_limits(x = 0)
```

## Publication-ready

```{r}
#| label: task-2

ggsave("rent_income_densities.pdf", width = 8, height = 4)
```

```{r}
#| label: cleanup
#| include: false

file.remove("rent_income_densities.pdf")
```

# Part 2: mapping America's partisan shift

## Fetching data

```{r}
#| label: task-3
#| eval: false

pres = read_csv("data/countypres_2000-2024.csv")

glimpse(pres)
count(pres, year)
```

```{r}
#| label: task-3-real
#| echo: false

pres = read_tsv("../files/countypres_2000-2024.tab") |>
  select(year, state, state_po, county_name, county_fips, office,
         candidate, party, candidatevotes, totalvotes, version, mode)

glimpse(pres)
count(pres, year)
```

## Fixing problems

```{r}
#| label: task-4

pres = pres |>
  filter(!is.na(party), !is.na(county_fips)) |>
  mutate(county_fips = str_pad(county_fips, width = 5, pad = "0"))

glimpse(pres)
```

First, `read_csv()` guessed `county_fips` as a *number* (`<dbl>`). Second, the 2024 bookkeeping rows (`TOTAL VOTES CAST`, `OVERVOTES`, …) have an empty `party`, which `read_csv()` reads in as `NA`. County votes are overcalculated with these rows in place. We also drop the handful of state-level rows with no `county_fips` at all.

## Republican vote share over time

```{r}
#| label: task-5

rep_votes = pres |>
  filter(party %in% c('DEMOCRAT', 'REPUBLICAN')) |>
  summarize(
    rep_share = sum(candidatevotes[party == "REPUBLICAN"]) /
      sum(candidatevotes),
    .by = c(year, county_fips)
  )

head(rep_votes, 3)

rep_wide = rep_votes |>
  pivot_wider(
    names_from = year,
    values_from = rep_share,
    names_prefix = "r"
    ) |>
  mutate(shift = r2024 - r2012)

head(rep_wide, 3)
```

## Visualizing county-level shifts

```{r}
#| label: task-6

ggplot(rep_wide, aes(x = r2012, y = r2024)) +
  geom_point(size = 0.5) +
  geom_abline() +
  labs(x = "Republican share, 2012", y = "Republican share, 2024",
       title = "County vote shares, 2024 vs. 2012") +
  theme_bw(base_size = 20)
```

## Mapping the shifts

```{r}
#| label: task-7-setup
#| eval: false
library(sf)

shapes = st_read("https://princeton-ddss.github.io/r-bootcamp/files/county_shapes.geojson")
```

```{r}
#| label: task-7-setup-real
#| include: false

library(sf)

shapes = st_read("../files/county_shapes.geojson", quiet = TRUE)
```

```{r}
#| label: task-7

shift_map = shapes |>
  left_join(rep_wide, join_by(GEOID == county_fips))

ggplot(shift_map) +
  geom_sf(aes(fill = shift), linewidth = 0.02) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    guide = guide_colorbar(
      ticks.colour = "black",  # Makes ticks stand out against light colors
      ticks.linewidth = 1      # Makes the tick marks thicker
      )
    ) +
  labs(fill = "Shift in R share, 2012 to 2024") +
  theme_void(base_size = 15) +
  theme(legend.position = 'top')
```

# Part 3: Putting it together

## Republican shifts and county-level income

```{r}
#| label: task-8

shift_map = shift_map |>
  left_join(
    counties |> select(GEOID, income),
    join_by(GEOID)
    )

ggplot(shift_map, aes(x = income, y = shift)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(
    x = "Median household income ($)",
    y = "Shift in Republican share,\n2012 - 2024",
    title = "Median HH Income vs.\nShift in tpvs, 2012 - 2024"
    ) +
  theme_minimal(base_size = 15) +
  scale_x_continuous(labels = label_dollar(scale_cut = cut_short_scale()))
```

## Thinking like a social scientist

This is a *county-level* (aggregate) relationship. Gelman et al.'s *Red State, Blue State* shows that aggregate and individual-level correlations _can_ vary in direction. We should not conclude from this plot that "less wealthy voters swung right and wealthier voters swung left." To examine this relationship properly, we need *individual-level* data: surveys with income and vote choice (ANES, CES, exit polls).

------------------------------------------------------------------------

*Election data: MIT Election Data and Science Lab, "County Presidential Election Returns 2000–2024", [doi:10.7910/DVN/VOQCHQ](https://doi.org/10.7910/DVN/VOQCHQ) (CC0).*
