---
title: "R Bootcamp Day 2"
subtitle: "Base R"
author:
  - name: Christopher T. Kenny
    affiliations:
      - name: Data-Driven Social Science
format:
  projector-typst:
    keep-typ: false
    backend: touying
    touying-theme: simple
theme: ddss.typ
fontsize: "14pt"
codefont: "Fira Code"
bullet-incremental: true
toc: true
handout: false
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    results: hold
---

# Today

## Goals

:::{.nonincremental}

- Familiarity with major components of "base R"
- Familiarity with functions, iteration, and vectorization
- Practice with using R interactively
- This afternoon: Quarto and tidyverse

:::

## Sources

::: {.nonincremental}

- Wickham, Çetinkaya-Rundel, and Grolemund. [R for Data Science, 2e, Ch. 25: Functions](https://r4ds.hadley.nz/functions.html), especially §25.1-25.2
- Wickham, Çetinkaya-Rundel, and Grolemund. [R for Data Science, 2e, Ch. 26: Iteration](https://r4ds.hadley.nz/iteration.html), especially §26.1
- Harvard IQSS. [Math Prefresher for Political Scientists, Ch. 9: Objects, Functions, Loops](https://iqss.github.io/prefresher/13_functions_obj_loops.html), especially §§9.4, 9.6, and 9.7
- R Core Team. [`lapply()` documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/lapply.html)

:::

```{r}
#| echo: false
states <- data.frame(
  state = state.name,
  region = state.region,
  population = state.x77[, 'Population'],
  income = state.x77[, 'Income'],
  hs_grad = state.x77[, 'HS Grad'],
  murder_rate = state.x77[, 'Murder'],
  life_expectancy = state.x77[, 'Life Exp']
)
```

# Functions

## What is a function?

Functions make code reusable

. . .

Function components:

- Arguments: the inputs to a function
- Body: the code to be run
- Output: what gets returned

. . .

Calling a function executes the code on your inputs

```r
result <- function_name(argument)
```

## Why write a function?

Writing repetitive code is (1) harder to change and (2) easier to break

. . .

The calculation is duplicated:

```{r}
income_z <- (states$income - mean(states$income, na.rm = TRUE)) /
  sd(states$income, na.rm = TRUE)

murder_z <- (states$murder_rate - mean(states$murder_rate, na.rm = TRUE)) /
  sd(states$murder_rate, na.rm = TRUE)
```

. . .

If you want to change one thing, you need to find every instance of the code

. . .

Miss a copy and the results disagree due to stale code

## Writing a function

We can define any function we want with `function()`

. . .

```{r}
z_score <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

income_z <- z_score(states$income)
murder_z <- z_score(states$murder_rate)
head(income_z)
```

. . .

Edit `z_score()` directly to fix every later call

## Arguments

- Arguments take input values
- Can include defaults (e.g., `arg = 3`)

. . .

```{r}
percentage <- function(count, total) {
  100 * count / total
}

percentage(48, 100) # matched by position
percentage(48, total = 100) # can match by name
```

. . .

- Named arguments make calls easier to read
- Defaults can always be overwritten when making a function call

## Return values

- Functions return one object which can contain many values
- R returns the last evaluated expression by default

. . .

```{r}
quick_summary <- function(x, na.rm = TRUE) {
  c(
    min = min(x, na.rm = na.rm),
    mean = mean(x, na.rm = na.rm),
    max = max(x, na.rm = na.rm)
  )
}

quick_summary(states$income)
```

## Local objects inside functions

Values defined within a function don't impact those outside

```{r}
mean_center <- function(x) {
  x_mean <- mean(x)
  x - x_mean
}

x_mean <- 100
mean_center(c(2, 4, 6))
x_mean
```

## Conditionals

Conditionals allow for forking paths based on logical values

. . .

```{r}
classify_change <- function(x) {
  if (x > 0) {
    'positive'
  } else if (x == 0) {
    'zero'
  } else {
    'negative'
  }
}

c(classify_change(5), classify_change(0), classify_change(-2))
```

## Vectorized conditionals

- `if` expects one `TRUE` or `FALSE`

. . .

```{r}
#| error: true
turnout_change <- c(-2, 0, 4)

if (turnout_change > 0) {
  'positive'
}
```

. . .

- Use `ifelse()` for element-by-element vector comparisons

. . .

```{r}
ifelse(turnout_change > 0, 'increase', 'no increase')
```

## Exercise 1: proportion above a cutoff

Write `prop_above()` to compute the proportion of values above a cutoff

```r
prop_above <- function(x, cutoff, na.rm = TRUE) {
  # your code here
}
```

. . .

It should handle this input:

```r
turnout <- c(0.71, 0.88, NA, 0.93, 0.65)
prop_above(turnout, cutoff = 0.80)
```

## A Solution

```{r}
prop_above <- function(x, cutoff, na.rm = TRUE) {
  mean(x > cutoff, na.rm = na.rm)
}

turnout <- c(0.71, 0.88, NA, 0.93, 0.65)
prop_above(turnout, cutoff = 0.80)
```

. . .

Why does this work?

. . .

```{r}
turnout > 0.80
```

R treats `TRUE` as 1 and `FALSE` as 0 when computing the mean

# Loops and iteration

## Vectorization

- R is vectorized: you can apply one operation to the full vector
- No iteration needed

. . .

```{r}
turnout <- c(0.52, 0.61, 0.68)

100 * turnout
```

. . .

Math is vectorized, as are most functions in R

. . .

What if I need to work on elements one-by-one?

## `for` loops

- Process one value at a time
- Repeat the body of the loop for each value

. . .

```{r}
states_of_interest <- c('California', 'Massachusetts', 'New Hampshire')

for (state in states_of_interest) {
  print(paste('Selected state:', state))
}
```

. . .

```{r}
state
```

## Saving results


:::{.nonincremental}

- Create a vector to store the outputs
- Use the index to fill one position on each iteration

:::

```{r}
columns <- states[c('income', 'hs_grad', 'murder_rate')]
column_means <- numeric(length(columns))
```

. . .

```{r}
for (i in seq_along(columns)) {
  column_means[i] <- mean(columns[[i]])
}
```

. . .

Give it names:

```{r}
names(column_means) <- names(columns)
column_means
```

## `while` loops

- Repeat while *something* remains `TRUE`
- Useful when the number of iterations is unknown
- The condition must eventually become `FALSE` or your computer will die

. . .

```{r}
daily_interviews <- c(18, 27, 24, 35, 28, 31)
interviews <- 0
day <- 0

while (interviews < 75) {
  day <- day + 1
  interviews <- interviews + daily_interviews[day]
}

c(days = day, interviews = interviews)
```

# The apply family

## `lapply()`: your new best friend

- Apply a function to each element of a vector, list, etc.
- Return one *list* element for each input element
- Preserve input names

. . .

```{r}
columns <- states[c('income', 'hs_grad', 'murder_rate')]
lapply(columns, mean)
```

. . .

Note: we pass the function as `mean`, not `mean()`

## Additional arguments

Arguments after the function are passed to every call

. . .

```{r}
trust_by_wave <- list(
  wave_1 = c(2, 3, NA, 4),
  wave_2 = c(3, NA, 4, 5)
)

lapply(trust_by_wave, mean, na.rm = TRUE)
```

. . .

### Warning! {.warning}

This is *not* advised, but is valid code and works!

## Anonymous functions

- Define a function where it is used once

. . .

```{r}
lapply(
  columns,
  function(x) c(mean = mean(x), sd = sd(x))
)
```

## Anonymous functions

::: {.nonincremental}

- A second way to write it

:::

```{r}
lapply(
  columns,
  \(x) c(mean = mean(x), sd = sd(x))
)
```

## Output types

- `lapply()` always returns a list
- `sapply()` simplifies when possible

. . .

```{r}
lapply(columns, mean)
sapply(columns, mean)
```

## Output types: A stricter choice

- `vapply()` requires a declared output type

. . .

```{r}
vapply(columns, mean, FUN.VALUE = numeric(1))
```

. . .

```{r}
#| error: true
vapply(columns, mean, FUN.VALUE = character(1))
```

## Split, apply, combine

- Split data into groups
- Apply one function to every group
- Combine the results

. . .

```{r}
states_by_region <- split(states, states$region)

mean_income <- lapply(
  states_by_region,
  \(data) mean(data$income)
)

unlist(mean_income)
```

## Iteration in base R

- These approaches return the same column means
- Choose the form that is easiest to explain and modify

. . .

```{r}
# explicit loop
out <- numeric(length(columns))
for (i in seq_along(columns)) {
  out[i] <- mean(columns[[i]])
}
names(out) <- names(columns)

out
sapply(columns, mean)
```

## Common mistakes

- Calling the function too early: `lapply(x, mean())`
- Not saving the result at each iteration
- Growing the output on every iteration
- Using `1:length(x)` when `x` may be empty (`1:0`)
- Expecting `lapply()` to return a vector
- Trying to do too much inside one loop

# Exercise

## Exercise 2: compare regions

Using `states`, compare regions

. . .

Create a function `region_summary()` that returns:

- Number of states
- Mean income
- Mean high school graduation rate
- Mean life expectancy

. . .

Apply it to the four regions

## Write and test the function

```r
region_summary <- function(data, na.rm = TRUE) {
  c(
    n = # ...,
    mean_income = # ...,
    mean_hs_grad = # ...,
    mean_life_expectancy = # ...
  )
}
```

. . .

Test on one group before adding iteration:

. . .

```r
northeast <- states[states$region == "Northeast", ]
region_summary(northeast)
```

## Split and apply

```r
states_by_region <- split(states, states$region)
result <- lapply(states_by_region, region_summary)

class(result)
str(result)
```

. . .

- What is the type of `result`?
- What is the type of each element?
- Which names were preserved?

## Missing values

- Add an `na.rm` argument
- Pass it to each summary function
- Test the function after adding an `NA`

. . .

```r
northeast$income[1] <- NA
region_summary(northeast)
```

Does the function still return four values?

## A Solution

```{r}
region_summary <- function(data, na.rm = TRUE) {
  c(
    n = nrow(data),
    mean_income = mean(data$income, na.rm = na.rm),
    mean_hs_grad = mean(data$hs_grad, na.rm = na.rm),
    mean_life_expectancy = mean(data$life_expectancy, na.rm = na.rm)
  )
}

states_by_region <- split(states, states$region)
result <- lapply(states_by_region, region_summary)
do.call(rbind, result)
```


# Wrap-up

## Summary

- Repeated calculation? Write a function
- Same calculation on a whole vector? Prefer vectorization to iteration
- Same function on each list element? Use `lapply()`
- Need to allow future iterations to depend on past ones? Use `for`
- Do not know the number of repetitions? Use `while`

. . .

```{=typst}
#v(2em)
```


### When to use iteration {.tip}

::: {.nonincremental}

- Start with working code
- Identify repetition
- Extract the repeated calculation

:::
