---
title: "Simulation exercise"
format:
  typst:
    mainfont: Georgia
    toc: false
    margin:
      x: 1.25cm
      y: 1.25cm
    number-sections: true
execute:
  message: false
  warning: false
---

Download this file's source ([day-3-sim-ex.qmd](https://raw.githubusercontent.com/princeton-ddss/r-bootcamp/main/notes/day-3-sim-ex.qmd)) and open it in RStudio, Positron, or VS Code. Work through the code and associated questions. **_Render as you go to ward off problems as early as possible._**\

**_For your own good, avoid AI tools for this task._** Use only `?function_name` and R documentation or the useful links in the slides.

Reminder: the `mean()` of a `TRUE`/`FALSE` vector gives you the **proportion that are `TRUE`**.

## Setup (given) {.unnumbered}

```{r}
#| label: setup

library(tidyverse)

set.seed(365)

counties <- left_join(
    read_csv("https://princeton-ddss.github.io/r-bootcamp/files/county_data.csv"),
    read_csv("https://princeton-ddss.github.io/r-bootcamp/files/county_population.csv"),
    join_by(GEOID == fips)
)
```

# Part 1: the birthday problem

In this section, we are going to work with the "Birthday Problem." *Suppose $k$ people gather in a room. Assuming birthdays are independent and equally likely on any day, what is the probability that at least two of them share a birthday?*

We won't derive a closed-form solution. Instead, we will simulate results.

## One room

Draw $k = 23$ birthdays and check whether any two of them match. Represent a birthday as a number from 1 to 365.

Hint: `duplicated()` and `anyDuplicated()` are useful here, but you don't *need* them.

```{r}
#| label: task-1

```

## Wrap it in a function

Write a function `shared_birthday(k)` that runs your last answer once for an arbitrary $k$ and returns a single `TRUE` or `FALSE`. Call it a few times to convince yourself it varies.

```{r}
#| label: task-2

```

## Many rooms

Use `replicate()` to run `shared_birthday(23)` 10,000 times and estimate the probability.

```{r}
#| label: task-3

```

## How many replications did you need?

Estimate the same probability using 10, 100, 1,000, 10,000, and 100,000 replications. How many replications do you need before the answer stops moving in the second decimal place?

```{r}
#| label: task-4

```

## Sweep across room sizes

Now vary the room. Estimate the probability for every $k$ from 2 to 60 and store the results in a data frame with columns `k` and `prob`. (For this, you can drop down to 1,000 replications if your laptop takes a while.)

```{r}
#| label: task-5

```

## Plot it

Plot the number of people in the room against the probability of a shared birthday. Roughly how many people do you need for the probability to pass 50%?

```{r}
#| label: task-6

```

## Check against the exact answer

There is a closed form. The probability that *nobody* shares a birthday among $k$ people is

$$\prod_{i=0}^{k-1} \frac{365 - i}{365}$$

so the probability at least two people do is one minus that. In R:

`prod((365 - 0:(k - 1)) / 365)`

Add the exact values to your plot as a second line to see how close 1,000 replications got you to the real answer.

```{r}
#| label: task-7

```

# Part 2: sampling and the bootstrap

For this part, **pretend the 3,135 counties are the population.** We'll draw samples from it to watch what sampling does.

## The population

A handful of counties are missing `pop`. Build a vector called `pop_all` holding the county populations with those dropped, and use it for the rest of this part.

Then, plot its distribution and report its mean and standard deviation. Note that it is nothing like a "bell curve."

```{r}
#| label: task-8

```

## Sampling distributions

For each sample size $n \in \{2, 5, 30, 100, 1000\}$, draw 10,000 samples of that size **with replacement** and record each sample's mean. Plot the five distributions of sample means, preferably using `facet_*()`.

(Sampling with replacement keeps each draw independent of the others, which makes the arithmetic in §2.3 exact.)

Hint: to facet, all five sets of means need to live in *one* data frame, with a column recording which sample size produced each mean. `map_dfr()` over the five sizes will build that, and so will a `for` loop.

Watch the shape change as $n$ grows. FYI -- It's converging to a distribution called the **normal distribution**.

```{r}
#| label: task-9

```

## The width of those distributions

For each $n$, compute the standard deviation of your 10,000 sample means. Then print a data frame with one row per sample size and four columns: the sample size, that simulated standard deviation, $\sigma / \sqrt{n}$, and the ratio of the last two. ($\sigma$ is the population standard deviation from §2.1.)

The ratio column is the comparison: if the two agree, it sits near 1.

Note the $\sqrt{\ }$: to halve your uncertainty you need *four* times the data.

```{r}
#| label: task-10

```

## The middle 95%

Take your 10,000 sample means for $n = 100$ and find the values that cut off the bottom 2.5% and the top 2.5%. `quantile()` does this.

FYI -- this is a **confidence interval**. The area in the tails is a **p-value**.

```{r}
#| label: task-11

```

## The bootstrap

Everything above relied on having the whole population, which let us draw as many fresh samples as we liked, at any size we liked. **You will never be in that position.** In real research you have one sample, and the population is gone.

Run this, and from here on pretend `my_counties` is all the data that exists:

```{r}
#| label: bootstrap-setup

my_counties <- counties |>
    filter(!is.na(income)) |>
    slice_sample(n = 50)
```

Report the mean `income` in `my_counties`. That is your estimate, and the only number you have.

Then, to see how much it would have wobbled had you drawn a different 50 counties: treat `my_counties` as if it were the population and resample from *it*, with replacement, **at size 50** — the size you actually collected. Do that 10,000 times, plot the result, and report its standard deviation and its middle 95%.

(How much an estimate wobbles depends on how many observations it is built from. So unlike §2.2, where varying $n$ was the whole point, here it is fixed by the data you have.)

Why bother: your estimate is a single number, and on its own it says nothing about how much to trust it. The middle 95% is that missing piece — it is the range of values for the mean income of *all* US counties that your 50 counties are consistent with, arrived at without ever looking at those other 3,000-odd counties. Draw a fresh 50 and repeat the whole procedure, and intervals built this way land on the truth about 19 times in 20.

```{r}
#| label: task-12

```
