---
title: "Simulation exercise: solutions"
date: 2026-09-01
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  message: false
  warning: false
---

This document includes only one possible solution. Yours may vary — and because
these are simulations, your *numbers* will vary a little even if your code is
identical to mine. That is the point.

## Setup (given) {.unnumbered}

```{r}
#| label: setup

library(tidyverse)

set.seed(365)

counties <- left_join(
    read_csv("https://princeton-ddss.github.io/r-bootcamp/files/county_data.csv"),
    read_csv("https://princeton-ddss.github.io/r-bootcamp/files/county_population.csv"),
    join_by(GEOID == fips)
)
```

# Part 1: the birthday problem

## One room

```{r}
#| label: task-1

room <- sample(1:365, size = 23, replace = TRUE)

anyDuplicated(room) > 0
```

`replace = TRUE` is not optional. Without it, no two people *can* share a
birthday — `sample()` would hand out 23 different days by construction, and the
answer would always be `FALSE`.

`anyDuplicated()` returns the index of the first duplicate, or `0` if there
isn't one, so comparing to zero gives a logical. `any(duplicated(room))` works
just as well.

## Wrap it in a function

```{r}
#| label: task-2

shared_birthday <- function(k) {
    birthdays <- sample(1:365, size = k, replace = TRUE)
    anyDuplicated(birthdays) > 0
}

replicate(5, shared_birthday(23))
```

## Many rooms

```{r}
#| label: task-3

rooms <- replicate(10000, shared_birthday(23))

mean(rooms)
```

About `r round(mean(rooms) * 100)`% — a coin flip. Most people guess something
closer to 6%.

## How many replications did you need?

```{r}
#| label: task-4

reps <- c(10, 100, 1000, 10000, 100000)

tibble(
    replications = reps,
    estimate = sapply(reps, function(r) mean(replicate(r, shared_birthday(23))))
)
```

Ten replications can't even give you two decimal places — the answer is a
multiple of 0.1. By 1,000 you have the first decimal reliably; by 10,000 the
second is roughly stable. Going to 100,000 changes almost nothing, which is
how you know you have enough.

Note that the *question* never changed: it was 23 people every time. All that
changed was how hard we looked.

## Sweep across room sizes

```{r}
#| label: task-5

ks <- 2:60

birthday_curve <- tibble(
    k = ks,
    prob = sapply(ks, function(k) mean(replicate(1000, shared_birthday(k))))
)

head(birthday_curve, 3)
```

## Plot it

```{r}
#| label: task-6

ggplot(birthday_curve, aes(x = k, y = prob)) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0.5, linetype = "dotted") +
    geom_vline(xintercept = 23, linetype = "dotted") +
    labs(
        x = "People in the room",
        y = "Pr(at least one shared birthday)",
        title = "The birthday problem, simulated"
    ) +
    theme_minimal(base_size = 15)

min(birthday_curve$k[birthday_curve$prob > 0.5])
```

Twenty-three. The curve is S-shaped: nearly flat for the first handful of
people, then it climbs startlingly fast, and by 60 people it is
indistinguishable from certainty.

## Check against the exact answer

```{r}
#| label: task-7

exact_prob <- function(k) 1 - prod((365 - 0:(k - 1)) / 365)

birthday_curve <- birthday_curve |>
    mutate(exact = sapply(k, exact_prob))

exact_prob(23)

birthday_curve |>
    pivot_longer(c(prob, exact), names_to = "source", values_to = "p") |>
    mutate(source = if_else(source == "prob", "Simulated", "Exact")) |>
    ggplot(aes(x = k, y = p, color = source)) +
    geom_line(linewidth = 1) +
    labs(
        x = "People in the room",
        y = "Pr(at least one shared birthday)",
        color = NULL
    ) +
    theme_minimal(base_size = 15) +
    theme(legend.position = "bottom")
```

The two curves are hard to tell apart. The largest gap anywhere on the
sweep is given by:

`round(max(abs(birthday_curve$prob - birthday_curve$exact)), 3)`

## In writing

The question sounds like "does anyone share *my* birthday?", which for 23
people is indeed about 6%. But the question asked is whether *any pair* shares
one, and 23 people make 253 pairs. The number of chances to match grows with
the square of the room, not with the room. People substitute the easier
question without noticing they've done it.

# Part 2: sampling and the bootstrap

## The population

```{r}
#| label: task-8

pop_all <- counties$pop[!is.na(counties$pop)]

length(pop_all)

ggplot(tibble(pop = pop_all), aes(x = pop)) +
    geom_histogram(bins = 60) +
    scale_x_continuous(
        labels = scales::label_number(scale_cut = scales::cut_short_scale())
    ) +
    labs(x = "County population", y = "Count") +
    theme_minimal(base_size = 15)

mean(pop_all)
sd(pop_all)
```

Nothing like a bell. Almost every county is jammed against zero, with a long thin
tail out to Los Angeles County at `r format(max(pop_all), big.mark = ",")`. The
mean (`r format(round(mean(pop_all)), big.mark = ",")`) is four times the median
(`r format(round(median(pop_all)), big.mark = ",")`), which is what extreme skew
does to an average.

## Sampling distributions

```{r}
#| label: task-9

sizes <- c(2, 5, 30, 100, 1000)

samp <- map_dfr(sizes, function(n) {
    tibble(
        n = n,
        xbar = replicate(10000, mean(sample(pop_all, size = n, replace = TRUE)))
    )
})

samp |>
    mutate(n = factor(n, levels = sizes, labels = paste("n =", sizes))) |>
    ggplot(aes(x = xbar)) +
    geom_histogram(bins = 50) +
    facet_wrap(~n, scales = "free", nrow = 1) +
    scale_x_continuous(
        labels = scales::label_number(scale_cut = scales::cut_short_scale())
    ) +
    labs(x = "Sample mean of county population", y = "Count") +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The x-axes are **free**, so read the axis labels rather than the apparent widths.
At `n = 2` the distribution still looks like the population — a spike on the left
with a tail. By `n = 100` it is nearly symmetric, and at `n = 1000` it is a clean
bell centred on the population mean.

## The width of those distributions

```{r}
#| label: task-10

samp |>
    summarize(sim_sd = sd(xbar), .by = n) |>
    mutate(
        theory = sd(pop_all) / sqrt(n),
        ratio = sim_sd / theory
    )
```

Every simulated spread lands within about 5% of $\sigma/\sqrt{n}$, and the agreement
tightens sharply as `n` grows — by `n = 1000` it is within a fraction of a percent.
(The scatter at `n = 2` and `n = 5`, in both directions, is not an error in the
formula. Estimating a standard deviation from a distribution with a tail this heavy
is itself noisy, and 10,000 replications is not many for that job.)

Doubling your sample size does not halve your uncertainty; it divides it by
$\sqrt{2} \approx 1.41$. Halving it takes four times the data, and cutting it by a
factor of 10 takes a hundred times.

## The middle 95%

```{r}
#| label: task-11

quantile(samp$xbar[samp$n == 100], c(0.025, 0.975))
```

95% of samples of 100 counties give a mean between those two numbers — so a sample
landing outside happens one time in twenty by chance alone, with nothing wrong.

The interval is not centred on the population mean
(`r format(round(mean(pop_all)), big.mark = ",")` sits left of the middle); the
skew is still faintly visible even after averaging 100 counties.

## The bootstrap

```{r}
#| label: bootstrap-setup

my_counties <- counties |>
    filter(!is.na(income)) |>
    slice_sample(n = 50)
```

```{r}
#| label: task-12

mean(my_counties$income)

boot50 <- replicate(
    10000,
    mean(sample(my_counties$income, size = 50, replace = TRUE))
)

ggplot(tibble(m = boot50), aes(x = m)) +
    geom_histogram(bins = 40) +
    labs(x = "Bootstrap mean income", y = "Count") +
    theme_minimal(base_size = 15)

sd(boot50)
quantile(boot50, c(0.025, 0.975))
```

That bell again — from 50 numbers, with no assumption about where they came from.

`replace = TRUE` is doing all the work. Without it, drawing 50 from 50 returns the
same 50 values in a new order, and a permutation has the same mean as the thing it
permutes:

```{r}
#| label: task-12b

mean(sample(my_counties$income, size = 50))
mean(sample(my_counties$income, size = 50))
```

Replacement is what lets some counties appear twice and others not at all, which is
what imitates the luck of the draw.

Resampling at size 10 instead would give a spread about $\sqrt{50/10} \approx 2.24$
times larger. That number is perfectly correct — it just describes a **10-county
study**, and you ran a 50-county one. Report it and you have overstated your own
uncertainty by a factor of two, having collected the data to do better.

Earlier in this part we had the population, so `n` was ours to choose and varying it
was the point. In a bootstrap the population is gone and `n` is a fact about the
data you have.

For real work, `boot::boot()` and `rsample::bootstraps()` do this plus the bias
corrections and interval methods you would otherwise have to look up.

------------------------------------------------------------------------

*County data: US Census Bureau, 2023 five-year American Community Survey.*
